manuscript_outline

### 2025-07-23 docs/manuscript/paper.Rmd

To address these limitations in fishes, @sparks2019 introduced the effective value model—an approach that captures the non-linear, temperature-dependent nature of development by integrating experimental data into a daily developmental rate function. Building on that work, the R package **hatchR** [@sparks2025] was developed to operationalize the effective value framework and make it accessible to researchers and managers. **hatchR** uses temperature-performance curves generated from lab-based thermal rearing trials to estimate the proportion of development achieved each day under observed temperature conditions. These daily contributions are summed until a cumulative threshold is reached (typically 1.0), marking the completion of a developmental stage. The package includes both an R-based workflow and a user-friendly Shiny web interface to facilitate broad adoption across research and management communities (<https://bmait101.github.io/hatchR/index.html>).

While **hatchR** was originally developed for coldwater fishes, the approach is broadly applicable to any poikilothermic organism whose development conforms to a monotonic or unimodal temperature–rate relationship, including many amphibians, reptiles, and invertebrates. In this manuscript, we expand the taxonomic scope of **hatchR**, compiling thermal developmental data from the literature and demonstrating how effective value modeling can be applied to diverse taxa. We also present three case studies illustrating how **hatchR** can inform ecological, evolutionary, and management questions—ranging from understanding thermal adaptation to forecasting hatch timing in conservation planning. These applications are increasingly relevant as ecosystems, especially aquatic ones, face intensifying and interacting stressors, including variable climates, habitat alteration, and invasive species, all of which can disrupt developmental timing and population dynamics [@craig2017; @pinsky2019].

@@

Effective value models estimate developmental progress based on empirically derived relationships between temperature and development rate. These relationships are typically parameterized by raising organisms at multiple constant temperatures in laboratory settings, then fitting a non-linear function that captures the temperature-dependence of development. This function can then be reciprocated to produce daily “effective values”—units of development—that accumulate over time.

@@

Although originally developed for fishes, **hatchR** is a generalizeable and flexible tool for predicting phenology across a broad set of poikilothermic organisms. The effective value framework accommodates nonlinear thermal sensitivity and allows researchers to estimate the timing of developmental milestones under real-world, fluctuating temperature regimes with high accuracy and precision. In this manuscript, we demonstrate how power-law models can be fit using the `fit_model()` function, or defined externally and passed into the `predict_phenology()` function. For users without coding experience, a graphical user interface is also available via the **hatchR** Shiny app, expanding accessibility and encouraging broader adoption across disciplines.

@@

Although initially developed for fishes, **hatchR** is a broadly applicable tool for modeling temperature-dependent development in poikilotherms. It addresses key limitations of traditional degree-day approaches by accounting for non-linear thermal response curves and enabling predictions under fluctuating thermal conditions. As global temperatures rise and the phenology of organisms shifts in response, tools like **hatchR** will be increasingly valuable for anticipating the ecological and evolutionary consequences of variable climates. While we provide numerous species-level parameterizations and demonstrate a set of representative case studies, the full range of possible applications—across taxa, ecosystems, and educational settings—extends far beyond what we present here.

>>>>

To address these limitations in fishes, @sparks2019 introduced the effective value model—an approach that captures the non-linear, temperature-dependent nature of development by integrating experimental data into a daily developmental rate function. Building on that work, the R package **hatchR** [@sparks2025] was developed to operationalize the effective value framework and make it accessible to researchers and managers. **hatchR** uses temperature-performance curves generated from lab-based thermal rearing trials to estimate the proportion of development achieved each day under observed temperature conditions. These daily contributions are summed until a cumulative threshold is reached, marking the completion of a developmental stage. The package includes both an R-based workflow and a user-friendly Shiny web interface to facilitate broad adoption across research and management communities (<https://bmait101.github.io/hatchR/index.html>).

While **hatchR** was originally developed for coldwater fishes, the approach is broadly applicable to any poikilothermic organism whose development conforms to a monotonic or unimodal temperature–rate relationship, including many amphibians, reptiles, and invertebrates. In this manuscript, we expand the taxonomic scope of **hatchR**, compiling thermal developmental data from the literature and demonstrating how effective value models can be applied to diverse taxa. We also present three case studies illustrating how **hatchR** can inform ecological, evolutionary, and management questions—ranging from understanding thermal adaptation to forecasting hatch timing in conservation planning. These applications are increasingly relevant as ecosystems, especially aquatic ones, face intensifying and interacting stressors, including variable climates, habitat alteration, and invasive species, all of which can disrupt developmental timing and population dynamics [@craig2017; @pinsky2019].

@@

Effective value models estimate developmental progress based on empirically derived relationships between temperature and development rate. These relationships are typically parameterized by raising organisms at multiple constant temperatures in laboratory settings, then fitting a non-linear function that captures the temperature-dependence of development (see Table 1 for examples). This function can then be reciprocated to produce daily “effective values”—units of development—that accumulate over time [@sparks2019].

@@

**hatchR** is a generalizeable and flexible tool for predicting phenology across a broad set of poikilothermic organisms. The effective value framework accommodates nonlinear thermal sensitivity and allows researchers to estimate the timing of developmental milestones under real-world, fluctuating temperature regimes with high accuracy and precision. In this manuscript, we demonstrate how power-law models can be fit using the `fit_model()` function, or defined externally and passed into the `predict_phenology()` function. For users without coding experience, a graphical user interface is also available via the **hatchR** Shiny app, expanding accessibility and encouraging broader adoption across disciplines.

@@

**hatchR** is a broadly applicable tool for modeling temperature-dependent development in poikilotherms. It addresses key limitations of traditional degree-day approaches by accounting for non-linear thermal response curves and enabling predictions under fluctuating thermal conditions. As global temperatures rise and the phenology of organisms shifts in response, tools like **hatchR** will be increasingly valuable for anticipating the ecological and evolutionary consequences of variable climates. While we provide numerous species-level parameterizations and demonstrate a set of representative case studies, the full range of possible applications—across taxa, ecosystems, and educational settings—extends far beyond what we present here.
